---
title: "Lab 04 - La Quinta is Spanish for next to Denny's, Pt. 1"
author: "Lindsey Wilson"
date: "2/5/23"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox) 
```

```{r load-data, message = FALSE}
states <- read.csv("data/states.csv")
dennys <- dennys
laquinta <- laquinta
```

### Exercise 1

Below we have the dimensions of the Denny's dataset:

```{r summary-dennys}
message("The Denny's dataset contains ", nrow(dennys), " rows and ", ncol(dennys), " columns")
```


### Exercise 2

And here we have the dimensions of the La Quinta dataset:

```{r summary-laquinta}
message("The La Quinta dataset contains ", nrow(laquinta), " rows and ", ncol(laquinta), " columns")
```

### Exercise 3

Based on the websites given in the lab, it doesn't look like there are any Denny's locations outside the US.
However, it does look like there are a few La Quinta's outside the US: two in Canada, six in Mexico, two in China, one in New Zealand, three in Turkey, one in. the UAB, one in. Chile, and one in Colombia.

### Exercise 4

I can think of a few ways to figure out from the data whether or not there are Denny's/La Quinta locations in the US.

One quick way might be to filter out results that don't have a zip code (or something weird like '00000')

Another suggestion is to look up the highest/lowest longitude and latitude in the continental US and then filter out results that fall outside of the box defined by those coordinates. You could take a similar approach for Hawaii and Alaska to cover them as well.

### Exercise 5

The code below calculates the number Denny's outside the US by filtering out Denny's locations with a state code that isn't in the `states` dataset and then counting the number of rows left over:

```{r dennys-outside-US}
dennys_outside_US_count <- dennys %>%
  filter(!(state %in% states$abbreviation)) %>%
  nrow()

  message("There are ", dennys_outside_US_count, " Denny's locations outside the US")
```

### Exercise 6

Since we know there aren't any Denny's outside the US, we can go ahead and add a country variable to the Denny's dataset and set the value for all rows as "United States"

```{r add-country-variable-to-dennys}
dennys <- dennys %>%
  mutate(country = "United States")
```

### Exercise 7

We can take the same approach to find how many La Quinta locations are outside the US as well:

```{r laquinta-outside-US}
laquinta_outside_US_count <- laquinta %>%
  filter(!(state %in% states$abbreviation)) %>%
  nrow()

  message("There are ", laquinta_outside_US_count, " La Quinta locations outside the US")
```

And a little extra code gives us more information about those locations:

```{r laquinta-outside-US-locations}
laquinta_outside_US <- laquinta %>%
   filter(!(state %in% states$abbreviation))

laquinta_outside_US
```


### Exercise 8

After some Googling, I'm able to figure out which countries these locations are in based on their country codes:

Codes AG,  QR, CH, NL, VE,  PU, and SL are in Mexico;
Code ANT is in Colombia;
Codes ON and BC are in Canada;
and Code FM in is Honduras.

We can use this knowledge to add a country code to the La Quinta data as well, and then filter the data so that we're only working with observations from within the US:

```{r add-country-variable-to-la-quinta}
laquinta <- laquinta %>%
  mutate(country = case_when(
    state %in% state.abb     ~ "United States",
    state %in% c("AG", "QR", "CH",  "NL", "VE", "PU", "SL")  ~"Mexico",
    state %in% c("ON", "BC") ~ "Canada",
    state == "ANT"           ~ "Colombia",
    state == "FM"            ~ "Honduras" 
    ))

laquinta <- laquinta %>%
  filter(country == "United States")
```

### Exercise 9

It looks like California has the most Denny's locations (with 403), while Delaware has the fewest (just 1)

```{r dennys-counts-by-state}
dennys %>%
  count(state) %>%
  arrange(desc(n)) %>%
  print(n = 100)
```

For La Quinta, it looks like Texas has the most locations (237) and Maine has the fewest (1)

```{r laquinta-counts-by-state}
laquinta %>%
  count(state) %>%
  arrange(desc(n)) %>%
  print(n=100)
```

### Exercise 10

To figure out how many locations of Denny's and La Quinta are in each state per 1000 miles, we have to combine the counts of each location per state with the `states` data frame.
We can then create a new variable within each merged data frame called "LocationsPerArea" that divides the location count per state by (area/1000).
Finally, we can sort each data frame in descending order by this new variable.

Below, we can see that DC has the highest concentration of Denny's (29.2 per 1000 square miles), and Rhode Island has the highest concentration of La Quinta (1.29 per 1000 square miles)

```{r locations-per-thousand miles}
dennys_joined <- dennys %>%
  count(state) %>%
  inner_join(states, by = c("state" = "abbreviation")) %>%
  mutate(LocationsPerArea = n/(area/1000)) %>%
  arrange(desc(LocationsPerArea))
dennys_joined

laquinta_joined <- laquinta %>%
  count(state) %>%
  inner_join(states, by = c("state" = "abbreviation")) %>%
  mutate(LocationsPerArea = n/(area/1000)) %>%
  arrange(desc(LocationsPerArea))
laquinta_joined
```
### Exercise 11

To compare the locations of Denny's and La Quinta, we need to combine the two datasets together.
The code to do that is given below:

```{r merge-dn-lq}
dennys <- dennys %>%
  mutate(establishment = "Denny's")
laquinta <- laquinta %>%
  mutate(establishment = "La Quinta")

dn_lq <- bind_rows(dennys, laquinta)
```

And now that the datasets are joined, we can make a scatterplot of latitude vs. longitude for the two establishments.
This should allow us to literally see where they are in relation to one another:

```{r location-scatterplot-US}
ggplot(dn_lq, mapping = aes(x = longitude,
                            y = latitude,
                            color = establishment)) +
  geom_point()
```

Let's filter this to look just at the data for North Carolina:

```{r location-scatterplot-NC}
dn_lq %>%
  filter(state == "NC") %>%
ggplot(mapping = aes(x = longitude,
                     y = latitude,
                     color = establishment)) +
  geom_point(alpha = 0.7)
```

I'd say this provides some support for Mitch Hedberg's joke.
Not every Denny's is right next to a La Quinta, but there are quite a few with a La Quinta very close by.

### Exercise 12

Now let's do the same analysis for Texas:

```{r location-scatterplot-TX}
dn_lq %>%
  filter(state == "TX") %>%
ggplot(mapping = aes(x = longitude,
                     y = latitude,
                     color = establishment)) +
  geom_point(alpha = 0.5)
```
This is a lot more visually striking.
It looks like there are a few population centers in Texas, and those are rich with both Denny's and La Quintas.
And even for those outside of the big clusters, the proximity is a little uncanny.

Looks like Mitch was onto something!

